---
title: "Cell Type Deconvolution"
output: html_notebook
---

```{r, warning=FALSE, message=FALSE, include=FALSE}
library(tidyverse)
library(patchwork)
library(Metrics)
```

## CIBERSORTx Assessment
Whether aggregating by (cell line, day) or by (cell line, sample), RMSE does not appear to be significantly impacted, and cell type deconvolution appears reasonably accurate
```{r, , warning=FALSE, message=FALSE, echo=FALSE}
true.day <- read_tsv("results/cibersort/pseudobulk_fracs.indday.test.true.tsv")
inferred.day <- read_tsv("results/cibersort/pseudobulk_fracs.indday.test.inferred.txt") %>% 
  select(!c(`P-value`, Correlation, RMSE)) %>%
  rename(sample=Mixture) %>%
  relocate(colnames(true.day))
stopifnot(all.equal(true.day$sample, inferred.day$sample))
stopifnot(all.equal(colnames(true.day), colnames(inferred.day)))
cell.types <- colnames(true.day)[-c(1)]
rmse.day <- bind_rows(lapply(cell.types, function(ct){tibble("type"=ct, "rmse"=rmse(true.day[[ct]], inferred.day[[ct]]))}))

true.col <- read_tsv("results/cibersort/pseudobulk_fracs.indcol.test.true.tsv")
inferred.col <- read_tsv("results/cibersort/pseudobulk_fracs.indcol.test.inferred.txt") %>% 
  select(!c(`P-value`, Correlation, RMSE)) %>%
  rename(sample=Mixture) %>%
  relocate(colnames(true.col))
stopifnot(all.equal(true.col$sample, inferred.col$sample))
stopifnot(all.equal(colnames(true.col), colnames(inferred.col)))
cell.types <- colnames(true.col)[-c(1)]
rmse.col <- bind_rows(lapply(cell.types, function(ct){tibble("type"=ct, "rmse"=rmse(true.col[[ct]], inferred.col[[ct]]))}))

rmse.comp <- rmse.day %>%
  rename(day=rmse) %>%
  inner_join(rmse.col, by="type") %>%
  rename(col=rmse) %>%
  gather(!type, key="aggregation", value="rmse")
ggplot(rmse.comp, aes(x=type, y=rmse, fill=aggregation)) +
  geom_bar(position="dodge", stat="identity") +
  xlab("Cell Type") + 
  ylab("RMSE")
```

We can also visually see that the true and inferred cell type proportions as they change over time are similar
```{r, warning=FALSE, message=FALSE, echo=FALSE}
cell.types <- c("iPSC", "meso", "EMT", "cardiomes", "prog", "CM", "EPDC")
true.day <- true.day %>% 
  gather(!sample, key="type", value="frac") %>% 
  mutate(day=str_extract(sample, "[^_]+$")) %>%
  mutate(ind=str_extract(sample, "[^_]+")) %>%
  mutate(type=factor(type, levels=cell.types)) %>%
  mutate(day=as.numeric(day)) %>%
  mutate(ind=factor(ind)) %>%
  mutate(deconv="true")
inds.3 <- sample(unique(true.day$ind), 3)
true.day <- filter(true.day, ind %in% inds.3)

inferred.day <- inferred.day %>% 
  gather(!sample, key="type", value="frac") %>% 
  mutate(day=str_extract(sample, "[^_]+$")) %>%
  mutate(ind=str_extract(sample, "[^_]+")) %>%
  mutate(type=factor(type, levels=cell.types)) %>%
  mutate(day=as.numeric(day)) %>%
  mutate(ind=factor(ind)) %>%
  mutate(deconv="inferred") %>%
  filter(ind %in% inds.3)

p <- bind_rows(true.day, inferred.day)

ggplot(p, aes(x=day, y=frac, fill=type)) +
  geom_area(alpha=0.6 , size=1, colour="black") +
  facet_grid(rows=vars(ind), cols=vars(deconv))+
  theme_classic() + theme(axis.ticks.y = element_blank(), axis.text.y=element_blank())

```


## Bulk Deconvolution
Bulk deconvolution suggests that while cell lines do not consistently follow the same differentiation trajectory between the two experiments, there does seem to be a notable trend of 
```{r, fig.width=15, fig.height=10, echo=FALSE}
bulk <- read_tsv("results/cibersort/bulk.inferred.tsv") %>% 
  select(!c(`P-value`, Correlation, RMSE)) %>%
  rename(sample=Mixture) %>%
  gather(!sample, key="type", value="frac") %>% 
  mutate(day=str_extract(sample, "[^_]+$")) %>%
  mutate(ind=str_extract(sample, "[^_]+")) %>%
  mutate(type=factor(type, levels=cell.types)) %>%
  mutate(day=as.numeric(day)) %>%
  mutate(ind=factor(ind)) %>%
  mutate(exp="bulk")

pseudobulk <- read_tsv("results/cibersort/pseudobulk_fracs.indday.full.true.tsv") %>% 
  gather(!sample, key="type", value="frac") %>% 
  mutate(day=str_extract(sample, "[^_]+$")) %>%
  mutate(ind=str_extract(sample, "[^_]+")) %>%
  mutate(type=factor(type, levels=cell.types)) %>%
  mutate(day=as.numeric(day)) %>%
  mutate(ind=factor(ind)) %>%
  mutate(exp="pseudobulk")

p <- bind_rows(bulk, pseudobulk)

ggplot(p, aes(x=day, y=frac, fill=type)) +
  geom_area(alpha=0.6 , size=1, colour="black") +
  facet_grid(rows=vars(ind), cols=vars(exp)) +
  theme_classic() + theme(axis.ticks.y = element_blank(), axis.text.y=element_blank())
```
